---
title: "Low-complexity reefs as refugia for Mediterranean marine forests: analyses and figures"
author: "Sarah Ohayon"
date: "October 2026"
output: html_document
---

Analyses for the revision of MERE-D-26-01577. The models here replace the ones in
`UVC_Data.Rmd` and `2025_07_27.Rmd`: site is now a random effect, depth is included as
a covariate, and the grazing model accounts for transect and space.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
# paths are relative to this file (R/); data in ../data, figures written to ../figures
data_dir <- file.path("..", "data")
fig_dir  <- file.path("..", "figures")
dir.create(fig_dir, showWarnings = FALSE)
set.seed(42)
```

```{r libraries}
library(tidyverse)
library(mgcv)
library(ggpubr)
library(patchwork)
```

```{r plot style}
# ---- Colours ----
algae_cols <- c(
  "Gongolaria" = "#F9A602",
  "Sargassum"  = "#B1560f"
)

grazing_cols <- c(
  "Intact"          = "#009E73",
  "Partially eaten" = "#E69F00",
  "Eaten"           = "#b5651d"
)

treatment_cols <- c(
  "Control"      = "#598BAF",
  "Translocated" = "#29AB87"
)

# ---- Theme ----
# the theme I use for the paper figures, kept in one place so all the panels match
theme_paper <- theme_bw(base_size = 16) +
  theme(
    axis.title.x = element_text(margin = margin(t = 12)),
    axis.title.y = element_text(margin = margin(r = 12)),
    axis.text.x  = element_text(size = 14, color = "black"),
    axis.text.y  = element_text(size = 14, color = "black"),
    axis.line    = element_line(color = "black", linewidth = 0.5),
    axis.ticks   = element_line(color = "black", linewidth = 0.5),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(color = "grey85", linewidth = 0.5),
    legend.position  = "right"
  )

# the stats label that goes in the corner of each model panel
gam_label <- function(model, term = 1) {
  st    <- summary(model)$s.table
  p     <- st[term, "p-value"]
  p_str <- if (p < 0.001) "p < 0.001" else paste0("p = ", formatC(p, 3, format = "f"))
  stat  <- if ("F" %in% colnames(st)) "F" else "χ²"
  paste0(
    "edf = ",  formatC(st[term, "edf"], 2, format = "f"),
    "\n", stat, " = ", formatC(st[term, ncol(st) - 1], 1, format = "f"),
    "\n", p_str,
    "\ndev. explained = ", formatC(100 * summary(model)$dev.expl, 1, format = "f"), "%"
  )
}

# put the legend inside the panel instead of stealing width beside it.
# ggplot2 renamed the argument at 3.5.0, so handle both.
legend_inside <- function(x, y) {
  box <- list(
    legend.background = element_rect(fill = alpha("white", 0.7), color = NA),
    legend.key        = element_rect(fill = NA, color = NA),
    legend.title      = element_text(size = 13),
    legend.text       = element_text(size = 12)
  )
  if (utils::packageVersion("ggplot2") >= "3.5.0") {
    do.call(theme, c(list(legend.position = "inside",
                          legend.position.inside = c(x, y)), box))
  } else {
    do.call(theme, c(list(legend.position = c(x, y)), box))
  }
}
```

```{r read data}
fish_algae_data <- read.csv(file.path(data_dir, "fish_algae_data.csv"))

Algea_transect_2025_04_19 <- read.csv(file.path(data_dir, "2025_04_19_Algae_transect.csv"))
Algea_transect_2025_04_29 <- read.csv(file.path(data_dir, "2025_04_29_Algae_transect.csv"))
complexity_2025_04_29     <- read.csv(file.path(data_dir, "2025_04_29_complexity_layer.csv"))

# distance from complex reef, sampled from the roughness layer in QGIS/python
# (see distance_from_complex_reef.py)
grazing_distance <- read.csv(file.path(data_dir, "grazing_distance_to_reef.csv"))

seascape_experiments <- read.csv(file.path(data_dir, "seascape_experiments_data.csv"))
maxn_experiments     <- read.csv(file.path(data_dir, "MaxN_seascape_experiments.csv"))
```


# 0. Figure 5 - macroalgal cover across complex reef edges (2010 surveys)

Percent cover of turf and canopy-forming macroalgae along distance gradients from the edge
of structurally complex reef at Carmel Head and Achziv. Weighted GAMs (weights = 1/SE^2)
are fitted per site.

```{r figure 5}
cover_2010 <- read.csv(file.path(data_dir, "shikmona_achziv_2010.csv"), check.names = FALSE) %>%
  rename_with(str_trim) %>%
  mutate(Site = recode(Site, "Shikmona" = "Carmel Head"),
         Site = factor(Site, levels = c("Carmel Head", "Achziv")),
         Algae_type = factor(Algae_type, levels = c("Turf", "Canopy")),
         Distance_num = as.numeric(Distance),
         SE_safe = ifelse(SE == 0, min(SE[SE > 0], na.rm = TRUE), SE),   # avoid infinite weights
         weight = 1 / SE_safe^2)

cover_cols <- c("Turf" = "#117799", "Canopy" = "#cc8000")

cover_panel <- function(site) {
  d   <- filter(cover_2010, Site == site)
  m   <- gam(Percent_cover ~ Algae_type + s(Distance_num, by = Algae_type, k = 5),
             data = d, weights = weight, method = "REML")
  nd  <- expand.grid(Distance_num = seq(min(d$Distance_num), max(d$Distance_num), length.out = 200),
                     Algae_type = levels(d$Algae_type))
  pr  <- predict(m, newdata = nd, se.fit = TRUE)
  nd  <- mutate(nd, fit = pr$fit, lwr = fit - 1.96 * pr$se.fit, upr = fit + 1.96 * pr$se.fit,
                Site = site)
  ggplot() +
    geom_ribbon(data = nd, aes(Distance_num, ymin = lwr, ymax = upr, fill = Algae_type),
                alpha = 0.30, colour = NA) +
    geom_line(data = nd, aes(Distance_num, fit, colour = Algae_type), linewidth = 1.4) +
    geom_point(data = d, aes(Distance_num, Percent_cover, colour = Algae_type), size = 2.8, alpha = 0.9) +
    facet_wrap(~ Site) +
    scale_x_continuous(breaks = sort(unique(d$Distance_num))) +
    scale_colour_manual(values = cover_cols) + scale_fill_manual(values = cover_cols) +
    scale_y_continuous(limits = c(-10, 115), expand = expansion(mult = c(0.02, 0.05))) +
    labs(x = "Distance from structurally complex reef edge (m)", y = "Percent cover (%)",
         colour = "Algae type", fill = "Algae type") +
    theme_bw(base_size = 16) +
    theme(panel.grid = element_blank(),
          axis.line = element_line(colour = "black", linewidth = 0.5),
          axis.ticks = element_line(colour = "black", linewidth = 0.5),
          strip.background = element_rect(fill = "grey95", colour = "black"),
          strip.text = element_text(face = "bold"),
          legend.position = "bottom")
}

fig5 <- (cover_panel("Carmel Head") / cover_panel("Achziv")) +
  plot_layout(guides = "collect") + plot_annotation(tag_levels = "a") &
  theme(legend.position = "bottom", plot.tag = element_text(face = "bold"))
fig5
ggsave(file.path(fig_dir, "Fig5_cover_distance.png"), fig5, width = 8, height = 8, dpi = 400)
```

# 1. Fish community and macroalgae along the complexity gradient

Reviewers asked for the transects to be treated as nested within sites and repeated over
seasons, and for depth to be tested as a covariate. Site is added as a random effect and
sampling period as a fixed factor with three levels.

```{r prepare uvc data}
fish_algae_data <- fish_algae_data %>%
  mutate(site_name = as.factor(site_name),
         season    = as.factor(season),
         period    = as.factor(paste(season, year, sep = "_")),
         log_biomass   = log10(Biomass + 0.01),
         log_abundance = log10(Abundance + 1))

table(fish_algae_data$period)
table(fish_algae_data$habitat)
```

```{r biomass models}
# published model - complexity only
gam_biomass_published <- gam(log_biomass ~ s(mean_complexity, bs = "tp", k = 3),
                             data = fish_algae_data, family = "gaussian", method = "REML")

# with site as a random effect and sampling period
gam_biomass_site <- gam(log_biomass ~ s(mean_complexity, bs = "tp", k = 5) + period +
                          s(site_name, bs = "re"),
                        data = fish_algae_data, family = "gaussian", method = "REML")

# adding depth
gam_biomass_depth <- gam(log_biomass ~ s(mean_complexity, bs = "tp", k = 5) + s(depth, bs = "tp", k = 5) +
                           period + s(site_name, bs = "re"),
                         data = fish_algae_data, family = "gaussian", method = "REML")

gam_biomass_null <- gam(log_biomass ~ period + s(site_name, bs = "re"),
                        data = fish_algae_data, family = "gaussian", method = "REML")

summary(gam_biomass_depth)
AIC(gam_biomass_published, gam_biomass_site, gam_biomass_depth, gam_biomass_null)

# is the complexity effect independent of depth?
concurvity(gam_biomass_depth, full = FALSE)$estimate
```

```{r abundance and richness models}
gam_abundance_depth <- gam(log_abundance ~ s(mean_complexity, bs = "tp", k = 5) + s(depth, bs = "tp", k = 5) +
                             period + s(site_name, bs = "re"),
                           data = fish_algae_data, family = "gaussian", method = "REML")
gam_abundance_null  <- gam(log_abundance ~ period + s(site_name, bs = "re"),
                           data = fish_algae_data, family = "gaussian", method = "REML")

gam_richness_depth <- gam(Richness ~ s(mean_complexity, bs = "tp", k = 5) + s(depth, bs = "tp", k = 5) +
                            period + s(site_name, bs = "re"),
                          data = fish_algae_data, family = "gaussian", method = "REML")
gam_richness_null  <- gam(Richness ~ period + s(site_name, bs = "re"),
                          data = fish_algae_data, family = "gaussian", method = "REML")

summary(gam_abundance_depth)
summary(gam_richness_depth)
AIC(gam_abundance_depth, gam_abundance_null)
AIC(gam_richness_depth, gam_richness_null)
```

```{r algae height model}
gam_algae_depth <- gam(mean_algae_height ~ s(mean_complexity, bs = "tp", k = 5) + s(depth, bs = "tp", k = 5) +
                         period + s(site_name, bs = "re"),
                       data = fish_algae_data, family = "gaussian", method = "REML")
gam_algae_null  <- gam(mean_algae_height ~ period + s(site_name, bs = "re"),
                       data = fish_algae_data, family = "gaussian", method = "REML")

summary(gam_algae_depth)
AIC(gam_algae_depth, gam_algae_null)
concurvity(gam_algae_depth, full = FALSE)$estimate
```

```{r habitat summary}
# means per habitat, for the results text
uvc_summary <- fish_algae_data %>%
  group_by(habitat) %>%
  summarise(n = n(),
            complexity = mean(mean_complexity),
            biomass = mean(Biomass), biomass_sd = sd(Biomass),
            abundance = mean(Abundance), abundance_sd = sd(Abundance),
            richness = mean(Richness), richness_sd = sd(Richness),
            algae_height = mean(mean_algae_height), algae_height_sd = sd(mean_algae_height))

as.data.frame(uvc_summary)
```

```{r figure 7}
# partial effect of complexity, with depth at its mean and the site effect left out
gam_effect <- function(model, data, yvar, ylab, corner = c("left", "right")) {
  # which top corner the stats label sits in - panel d is hidden by the points
  # and the trend line on the left, so it takes corner = "right"
  corner <- match.arg(corner)
  lab_x  <- if (corner == "left") -Inf else Inf
  lab_h  <- if (corner == "left") -0.08 else 1.08
  nd <- data.frame(
    mean_complexity = seq(min(data$mean_complexity), max(data$mean_complexity), length.out = 200),
    depth     = mean(data$depth, na.rm = TRUE),
    period    = factor(levels(data$period)[1],    levels = levels(data$period)),
    site_name = factor(levels(data$site_name)[1], levels = levels(data$site_name))
  )
  pr <- predict(model, newdata = nd, se.fit = TRUE, exclude = "s(site_name)")
  nd$fit <- pr$fit
  nd$lwr <- pr$fit - 1.96 * pr$se.fit
  nd$upr <- pr$fit + 1.96 * pr$se.fit

  ggplot(data, aes(x = mean_complexity, y = .data[[yvar]])) +
    geom_point(size = 2, alpha = 0.7) +
    geom_ribbon(data = nd, aes(x = mean_complexity, ymin = lwr, ymax = upr),
                inherit.aes = FALSE, fill = "#598BAF", alpha = 0.18) +
    geom_line(data = nd, aes(x = mean_complexity, y = fit),
              inherit.aes = FALSE, color = "#598BAF", linewidth = 1.4) +
    annotate("text", x = lab_x, y = Inf, label = gam_label(model),
             hjust = lab_h, vjust = 1.15, size = 4.2, lineheight = 0.95, color = "black") +
    labs(x = "Visual complexity (0-5)", y = ylab) +
    theme_paper
}

p_biomass   <- gam_effect(gam_biomass_depth,   fish_algae_data, "log_biomass",
                          expression("Fish biomass (log"[10]*" kg/150 m"^2*")"))
p_abundance <- gam_effect(gam_abundance_depth, fish_algae_data, "log_abundance",
                          expression("Fish abundance (log"[10]*" ind./150 m"^2*")"))
p_richness  <- gam_effect(gam_richness_depth,  fish_algae_data, "Richness",
                          expression("Species richness (species/150 m"^2*")"))
p_algae     <- gam_effect(gam_algae_depth,     fish_algae_data, "mean_algae_height",
                          "Mean macroalgal height (cm)", corner = "right")

# (e) share of fish biomass made up by rabbitfish, high- vs low-complexity habitat
fish_raw <- read.csv(file.path(data_dir, "bat_galim_fish_2024_2025_analysis.csv"))
species_ab <- read.csv(file.path(data_dir, "species_a_b.csv"))
siganus_share <- fish_raw %>%
  left_join(species_ab, by = "species") %>%
  mutate(weight = new_a * length^new_b * abundance) %>%   # g; length-weight a, b from FishBase
  filter(habitat != "artificial.rock") %>%
  mutate(habitat = recode(habitat, "complex.rock" = "High complexity",
                          "macroalgal.beds" = "Low complexity"),
         group = ifelse(grepl("Siganus", species, ignore.case = TRUE), "Siganus spp.", "Other fish")) %>%
  group_by(habitat, group) %>%
  summarise(mean_weight_kg = mean(weight / 1000, na.rm = TRUE), .groups = "drop") %>%
  group_by(habitat) %>%
  mutate(percent = mean_weight_kg / sum(mean_weight_kg) * 100)
as.data.frame(siganus_share)

p_siganus <- ggplot(siganus_share, aes(x = "", y = percent, fill = group)) +
  geom_col(width = 1, colour = "white") +
  coord_polar(theta = "y") +
  facet_wrap(~ habitat) +
  geom_text(aes(label = paste0(group, "
 ", round(percent, 1), "%")),
            position = position_stack(vjust = 0.65), size = 4.5, lineheight = 0.9) +
  scale_fill_manual(values = c("Siganus spp." = "#2a8e9d", "Other fish" = "#bdbdbd")) +
  theme_void(base_size = 16) +
  theme(legend.position = "none", strip.text = element_text(size = 14, margin = margin(t = 4, b = 4)),
        panel.spacing = unit(1.2, "lines"))

fig7 <- ggarrange(ggarrange(p_biomass, p_abundance, p_richness, p_algae,
                            ncol = 2, nrow = 2, labels = c("a", "b", "c", "d")),
                  ggarrange(p_siganus, labels = "e"),
                  ncol = 1, heights = c(2, 0.9))
fig7
ggsave(file.path(fig_dir, "Fig7_fish_algae_complexity.png"), fig7, width = 12, height = 13.5, dpi = 400, bg = "white")

#ggsave("Fig7_fish_algae_complexity.pdf", fig7, width = 12, height = 9, dpi = 400)
```


# 2. Grazing status along the two algae transects

The two transects are joined as before, but the model now includes transect as a fixed
factor and a smooth of position to account for the fact that neighbouring plants along a
transect are not independent. Distance from the nearest structurally complex reef is
added as a predictor.

```{r prepare grazing data}
complexity_2025_04_29 <- complexity_2025_04_29 %>%
  dplyr::select(filename, rvalue_1)

Algea_transect_2025_04_29_full <- Algea_transect_2025_04_29 %>%
  left_join(complexity_2025_04_29, by = "filename") %>%
  rename(height_cm = algae_height) %>%
  mutate(sample = 2)

Algea_transect_2025_04_19 <- Algea_transect_2025_04_19 %>%
  rename(algae_type = algea_type) %>%
  dplyr::select(-HubDist) %>%
  mutate(sample = 1)

# a few algae_type entries have a trailing space - trim before filtering
algae_joined_trim <- function(x) mutate(x, across(c(algae_type, Eaten_whole), trimws))
Algea_transect_2025_04_19      <- algae_joined_trim(Algea_transect_2025_04_19)
Algea_transect_2025_04_29_full <- algae_joined_trim(Algea_transect_2025_04_29_full)

algae_joined <- bind_rows(Algea_transect_2025_04_19, Algea_transect_2025_04_29_full)

algae_joined <- algae_joined %>%
  left_join(grazing_distance %>% dplyr::select(filename, sample, itm_x, itm_y, dist_complex_reef),
            by = c("filename", "sample"))

grazing_status <- algae_joined %>%
  mutate(height_cm = suppressWarnings(as.numeric(height_cm))) %>%
  filter(!is.na(height_cm)) %>%
  filter(algae_type %in% c("Gongolaria", "Sargassum")) %>%
  mutate(date = as.Date(timestamp),
         Eaten_whole = case_when(
           Eaten_whole %in% c("Eaten", "eaten") ~ "Eaten",
           Eaten_whole %in% c("Whole", "whole") ~ "Intact",
           Eaten_whole %in% c("Semi", "semi")   ~ "Partially eaten")) %>%
  filter(!is.na(Eaten_whole)) %>%
  mutate(Eaten_whole = factor(Eaten_whole, levels = c("Intact", "Partially eaten", "Eaten")),
         grazing_num = as.numeric(Eaten_whole),
         sample = as.factor(sample))

table(grazing_status$sample, grazing_status$Eaten_whole)
```

```{r grazing models}
# as published - grazing status against substrate roughness at the plant
gam_grazing_rough <- gam(grazing_num ~ s(rvalue_1, bs = "tp", k = 5) + sample,
                         family = ocat(R = 3), data = grazing_status, method = "REML")

# against distance from the nearest structurally complex reef
gam_grazing_dist <- gam(grazing_num ~ s(dist_complex_reef, bs = "tp", k = 5) + sample,
                        family = ocat(R = 3), data = grazing_status, method = "REML")

gam_grazing_both <- gam(grazing_num ~ s(dist_complex_reef, bs = "tp", k = 5) +
                          s(rvalue_1, bs = "tp", k = 5) + sample,
                        family = ocat(R = 3), data = grazing_status, method = "REML")

gam_grazing_null <- gam(grazing_num ~ sample, family = ocat(R = 3),
                        data = grazing_status, method = "REML")

summary(gam_grazing_rough)
summary(gam_grazing_dist)
summary(gam_grazing_both)
AIC(gam_grazing_null, gam_grazing_rough, gam_grazing_dist, gam_grazing_both)
```

```{r spatial autocorrelation}
# plants were photographed about a metre apart, so neighbours are unlikely to be
# independent. Order them along each transect and check the residual correlation.
grazing_status <- grazing_status %>%
  group_by(sample) %>%
  mutate(order = rank(as.POSIXct(timestamp, format = "%Y-%m-%dT%H:%M:%S"), ties.method = "first")) %>%
  ungroup()

res <- residuals(gam_grazing_dist, type = "response")
for (sm in levels(grazing_status$sample)) {
  rr <- res[grazing_status$sample == sm][order(grazing_status$order[grazing_status$sample == sm])]
  cat(sprintf("transect %s: lag-1 residual autocorrelation r = %.3f (n = %d)\n",
              sm, cor(rr[-length(rr)], rr[-1]), length(rr)))
}

# thin the transects to increase the spacing between retained plants, and check that
# the distance effect survives
for (k in c(1, 2, 3, 5)) {
  sub <- grazing_status %>% filter(order %% k == 0 | k == 1)
  mk  <- gam(grazing_num ~ s(dist_complex_reef, bs = "tp", k = 5) + sample,
             family = ocat(R = 3), data = sub, method = "REML")
  st  <- summary(mk)$s.table
  cat(sprintf("every %d plant: n = %3d  edf = %.2f  chi2 = %5.1f  p = %.3g  dev.expl = %.1f%%\n",
              k, nrow(sub), st[1, "edf"], st[1, "Chi.sq"], st[1, "p-value"], 100 * summary(mk)$dev.expl))
}
```

```{r each transect separately}
# only two transects, so fit each one on its own and see whether the gradient is in both
for (s in levels(grazing_status$sample)) {
  sub <- grazing_status %>% filter(sample == s)
  m   <- gam(grazing_num ~ s(dist_complex_reef, bs = "tp", k = 5),
             family = ocat(R = 3), data = sub, method = "REML")
  st  <- summary(m)$s.table
  cat(sprintf("transect %s: n = %3d  range %.1f-%.1f m  edf = %.2f  chi2 = %5.1f  p = %.3g  dev.expl = %.1f%%\n",
              s, nrow(sub), min(sub$dist_complex_reef), max(sub$dist_complex_reef),
              st[1, "edf"], st[1, "Chi.sq"], st[1, "p-value"], 100 * summary(m)$dev.expl))
  print(round(tapply(sub$dist_complex_reef, sub$Eaten_whole, mean), 1))
}
```

```{r block bootstrap}
# with two transects and neighbouring plants a metre apart, the p-value from 256 rows is
# optimistic. Resample blocks of consecutive plants instead, so the bootstrap keeps the
# autocorrelation rather than breaking it, and report the effect as a probability difference.
pred_diff <- function(d) {
  m <- try(gam(grazing_num ~ s(dist_complex_reef, bs = "tp", k = 5) + sample,
               family = ocat(R = 3), data = d, method = "REML"), silent = TRUE)
  if (inherits(m, "try-error")) return(NA_real_)
  nd <- data.frame(dist_complex_reef = c(15, 45),
                   sample = factor(levels(grazing_status$sample)[1],
                                   levels = levels(grazing_status$sample)))
  p <- predict(m, newdata = nd, type = "response")
  p[1, 3] - p[2, 3]   # P(eaten) at 15 m minus P(eaten) at 45 m
}

grazing_ordered <- grazing_status %>% arrange(sample, order)
obs_diff <- pred_diff(grazing_ordered)

set.seed(42)
block_len <- 15
n_boot    <- 400
boot <- replicate(n_boot, {
  resampled <- lapply(levels(grazing_ordered$sample), function(s) {
    sub    <- grazing_ordered %>% filter(sample == s)
    n      <- nrow(sub)
    starts <- sample(seq_len(n - block_len + 1), ceiling(n / block_len), replace = TRUE)
    idx    <- unlist(lapply(starts, function(st) st:(st + block_len - 1)))[1:n]
    sub[idx, ]
  })
  suppressWarnings(pred_diff(bind_rows(resampled)))
})
boot <- boot[is.finite(boot)]

cat(sprintf("P(eaten | 15 m) - P(eaten | 45 m) = %.3f\n", obs_diff))
cat(sprintf("block bootstrap (%d plants per block, %d valid resamples): 95%% CI %.3f - %.3f\n",
            block_len, length(boot), quantile(boot, 0.025), quantile(boot, 0.975)))
cat(sprintf("resamples with a positive effect: %.0f%%\n", 100 * mean(boot > 0)))
```

```{r grazing summary}
grazing_summary <- grazing_status %>%
  group_by(Eaten_whole) %>%
  summarise(n = n(),
            dist_mean = mean(dist_complex_reef), dist_sd = sd(dist_complex_reef),
            dist_median = median(dist_complex_reef),
            rough_mean = mean(rvalue_1))

as.data.frame(grazing_summary)

# same thing within each transect, to show the gradient is not driven by one of them
grazing_status %>%
  group_by(sample, Eaten_whole) %>%
  summarise(n = n(), dist_median = median(dist_complex_reef), .groups = "drop") %>%
  as.data.frame()
```

```{r species specific height model}
# height of each photographed plant against the roughness under it, per species
height_species <- grazing_status %>%
  filter(!is.na(rvalue_1)) %>%
  mutate(algae_type = factor(algae_type))

gam_height_species <- gam(height_cm ~ algae_type + s(rvalue_1, by = algae_type, bs = "tp", k = 5) + sample,
                          data = height_species, method = "REML")
gam_height_species_null <- gam(height_cm ~ algae_type + sample,
                               data = height_species, method = "REML")

summary(gam_height_species)
AIC(gam_height_species_null, gam_height_species)
anova(gam_height_species_null, gam_height_species, test = "F")
```

```{r figure 8a}
# Figure 8 uses larger text than the other figures so it stays legible at page width
theme_big <- theme_bw(base_size = 22) +
  theme(axis.title.x = element_text(margin = margin(t = 12)),
        axis.title.y = element_text(margin = margin(r = 12)),
        axis.text  = element_text(size = 19, colour = "black"),
        axis.line  = element_line(colour = "black", linewidth = 0.5),
        axis.ticks = element_line(colour = "black", linewidth = 0.5),
        panel.grid.minor = element_blank(),
        panel.grid.major = element_line(colour = "grey85", linewidth = 0.5))
legend_big <- function(x, y) legend_inside(x, y) +
  theme(legend.title = element_text(size = 19), legend.text = element_text(size = 17))

# one grid per species, over that species' own roughness range - Sargassum only reaches
# about 1.7 degrees, so a shared grid drew its trend line far beyond where it has data
nd_sp <- do.call(rbind, lapply(levels(height_species$algae_type), function(sp) {
  rng <- range(height_species$rvalue_1[height_species$algae_type == sp], na.rm = TRUE)
  data.frame(
    rvalue_1   = seq(rng[1], rng[2], length.out = 200),
    algae_type = factor(sp, levels = levels(height_species$algae_type)),
    sample     = factor(levels(height_species$sample)[1], levels = levels(height_species$sample))
  )
}))
pr_sp <- predict(gam_height_species, newdata = nd_sp, se.fit = TRUE)
nd_sp$fit <- pr_sp$fit
nd_sp$lwr <- pr_sp$fit - 1.96 * pr_sp$se.fit
nd_sp$upr <- pr_sp$fit + 1.96 * pr_sp$se.fit

p8a <- ggplot(height_species, aes(x = rvalue_1, y = height_cm,
                                  color = algae_type, fill = algae_type)) +
  geom_point(size = 2.8, alpha = 0.5) +
  geom_ribbon(data = nd_sp, aes(y = fit, ymin = lwr, ymax = upr), alpha = 0.18, color = NA) +
  geom_line(data = nd_sp, aes(y = fit), linewidth = 1.9) +
  annotate("text", x = -Inf, y = Inf, label = gam_label(gam_height_species),
           hjust = -0.08, vjust = 1.15, size = 6, lineheight = 0.95, color = "black") +
  scale_color_manual(values = algae_cols, name = "Algae type") +
  scale_fill_manual(values  = algae_cols, name = "Algae type") +
  labs(x = "Substrate roughness (°)", y = "Algal height (cm)") +
  theme_big +
  legend_big(0.83, 0.62)

p8a

```

```{r figure 8b}
# probability of each grazing status along the distance gradient
# The last partially eaten plant sits at 39.0 m and the last eaten one at 31.1 m; beyond
# that every observation is intact, so the ordinal transition is extrapolated and the
# interval widens without bound. Draw the curve over the supported range instead.
dist_max <- 70

nd_d <- data.frame(
  dist_complex_reef = seq(min(grazing_status$dist_complex_reef, na.rm = TRUE),
                          min(max(grazing_status$dist_complex_reef, na.rm = TRUE), dist_max),
                          length.out = 200),
  sample = factor(levels(grazing_status$sample)[1], levels = levels(grazing_status$sample))
)

fit_p <- predict(gam_grazing_dist, newdata = nd_d, type = "response")

# 95% intervals by simulating from the coefficients, as before. ocat works on a latent
# scale, so the cut points are what turn a simulated draw into a probability per category
cuts <- c(-Inf, gam_grazing_dist$family$getTheta(TRUE), Inf)
set.seed(1)
Xp   <- predict(gam_grazing_dist, newdata = nd_d, type = "lpmatrix")
beta <- rmvn(1000, coef(gam_grazing_dist), vcov(gam_grazing_dist))
sim  <- Xp %*% t(beta)

pred_ci <- do.call(rbind, lapply(seq_along(levels(grazing_status$Eaten_whole)), function(i) {
  p_sim <- plogis(cuts[i + 1] - sim) - plogis(cuts[i] - sim)
  data.frame(
    dist_complex_reef = nd_d$dist_complex_reef,
    status = factor(levels(grazing_status$Eaten_whole)[i],
                    levels = levels(grazing_status$Eaten_whole)),
    fit = fit_p[, i],
    lwr = apply(p_sim, 1, quantile, 0.025),
    upr = apply(p_sim, 1, quantile, 0.975))
}))

p8b <- ggplot(pred_ci, aes(x = dist_complex_reef, y = fit, color = status, fill = status)) +
  geom_ribbon(aes(ymin = lwr, ymax = upr), alpha = 0.18, color = NA) +
  geom_line(linewidth = 1.9) +
  annotate("text", x = -Inf, y = Inf, label = gam_label(gam_grazing_dist),
           hjust = -0.08, vjust = 1.15, size = 6, lineheight = 0.95, color = "black") +
  # default ggplot hue palette, as in the originally submitted figure - keeps this panel
  # visually distinct from the algae colours in 8a
  scale_y_continuous(limits = c(0, 1), breaks = c(0, 0.2, 0.4, 0.6, 0.8, 1)) +
  labs(x = "Distance from complex reef (m)", y = "Probability of grazing status",
       color = "Grazing status", fill = "Grazing status") +
  theme_big +
  legend_big(0.79, 0.46)

p8b

```

```{r figure 8}
tag8 <- theme(plot.tag = element_text(size = 26, face = "bold"))
fig8 <- (p8a + labs(tag = "a") + tag8) | (p8b + labs(tag = "b") + tag8)
fig8
ggsave(file.path(fig_dir, "Fig8_height_and_grazing.png"), fig8, width = 16, height = 6.5, dpi = 400)
```


# 3. Translocation experiments

The camera station is the experimental unit. Grazing outcomes were scored from the video
at each station and are reported descriptively (controls were deployed on one date only
in the main experiment, so a formal test of grazing outcome is not appropriate).

```{r translocation}
# stations per date and treatment in the video data
seascape_experiments %>%
  filter(!is.na(Treatment) & Treatment != "") %>%
  distinct(Date, Treatment, Camera_ID) %>%
  count(Date, Treatment)

# grazing outcome per station, scored from the videos (as reported in the manuscript)
grazing_outcome <- tibble::tribble(
  ~experiment,               ~treatment,     ~stations, ~grazed,
  "Pilot, June 2024 (2 h)",  "Translocated",  4,         1,
  "Pilot, June 2024 (2 h)",  "Control",       2,         0,
  "Main, April 2025 (6 h)",  "Translocated",  5,         5,
  "Main, April 2025 (6 h)",  "Control",       2,         0)
grazing_outcome
```

```{r maxn by guild}
# functional groups, as in the original figure
fg_lookup <- c(
  "Caranx.crysos" = "Piscivores", "Dentex.dentex" = "Piscivores",
  "Diplodus.cervinus" = "Omnivores", "Diplodus.sargus" = "Omnivores",
  "Diplodus.vulgaris" = "Omnivores", "Epinephelus.costae" = "Piscivores",
  "Epinephelus.marginatus" = "Piscivores", "Fistularia.sp." = "Piscivores",
  "Lagocephalus.spadiceus" = "Piscivores", "Mugilidae" = "Omnivores",
  "Oblada.melanura" = "Omnivores", "Parupeneus.forsskali" = "Invertivores",
  "Pempheris.rhomboidea" = "Planktivores", "Siganus sp." = "Herbivores",
  "Siganus.sp." = "Herbivores", "Sparus.aurata" = "Omnivores",
  "Sphyraena.sp" = "Piscivores", "Symphodus tinca" = "Invertivores",
  "Thalassoma.pavo" = "Invertivores", "Torquigener.flavimaculosus" = "Invertivores")

maxn_experiments$functional_group <- fg_lookup[maxn_experiments$Species]
maxn_experiments$functional_group[is.na(maxn_experiments$functional_group)] <- "Other"

fg_obs <- maxn_experiments %>%
  filter(!is.na(Treatment), Treatment != "") %>%
  mutate(sample_id = paste(Date, Camera_ID, sep = "_")) %>%
  group_by(sample_id, Date, Treatment, functional_group) %>%
  summarise(total_maxn = sum(Max_n, na.rm = TRUE), .groups = "drop") %>%
  filter(!functional_group %in% c("Other", "Planktivores"))

# every deployment gets a value for every guild: a guild never seen on a video is MaxN = 0
deployments <- distinct(fg_obs, sample_id, Date, Treatment)
fg_sum <- tidyr::expand_grid(deployments,
                             functional_group = c("Herbivores", "Invertivores", "Omnivores", "Piscivores")) %>%
  left_join(fg_obs, by = c("sample_id", "Date", "Treatment", "functional_group")) %>%
  mutate(total_maxn = coalesce(total_maxn, 0))

count(deployments, Date, Treatment)

# exact two-sided permutation test: treatment labels are shuffled only among stations
# deployed on the same date (all 90 possible permutations)
perm_test <- function(d) {
  diff_means <- function(ctl) mean(d$total_maxn[-ctl]) - mean(d$total_maxn[ctl])
  obs  <- diff_means(which(d$Treatment == "Control"))
  byd  <- split(seq_len(nrow(d)), d$Date)
  opts <- lapply(byd, function(ix) combn(ix, sum(d$Treatment[ix] == "Control"), simplify = FALSE))
  grid <- expand.grid(lapply(opts, seq_along))
  stat <- apply(grid, 1, function(r) diff_means(unlist(Map(function(o, i) o[[i]], opts, r))))
  c(n_permutations = nrow(grid), p_value = mean(abs(stat) >= abs(obs) - 1e-9))
}
maxn_perm <- sapply(split(fg_sum, fg_sum$functional_group), perm_test)
round(maxn_perm, 3)
```

```{r figure 10}
guild_cols <- c(Herbivores = "#F8766D", Invertivores = "#7CAE00",
                Omnivores = "#00BFC4", Piscivores = "#C77CFF")
p_lab <- setNames(paste0("p = ", formatC(maxn_perm["p_value", ], digits = 2, format = "fg", flag = "#")),
                  colnames(maxn_perm))
p_lab["Herbivores"] <- paste0("p = ", formatC(maxn_perm["p_value", "Herbivores"], digits = 3, format = "f"))

theme_fig10 <- theme_bw(base_size = 22) +
  theme(axis.text = element_text(size = 19, colour = "black"),
        axis.ticks = element_line(colour = "black"), axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(), panel.grid.major = element_line(colour = "grey85"),
        strip.background = element_rect(fill = "grey95", colour = "grey20"),
        strip.text = element_text(size = 21, face = "bold", margin = margin(6, 0, 6, 0)),
        plot.tag = element_text(size = 28, face = "bold"), legend.position = "none")

guild_panel <- function(g, tag) {
  d <- fg_sum %>% filter(functional_group == g) %>%
    mutate(Treatment = factor(Treatment, levels = c("Control", "Translocated")))
  ggplot(d, aes(Treatment, total_maxn)) +
    geom_boxplot(fill = guild_cols[g], alpha = 0.85, width = 0.6, outlier.shape = NA,
                 linewidth = 0.7, colour = "black") +
    geom_jitter(width = 0.1, height = 0, size = 3.2, alpha = 0.65, colour = "grey10") +
    annotate("text", x = 1.5, y = max(d$total_maxn) * 1.12, label = p_lab[g], size = 7,
             fontface = if (g == "Herbivores") "bold" else "plain") +
    facet_wrap(~ functional_group) +
    scale_y_continuous(expand = expansion(mult = c(0.04, 0.12))) +
    labs(x = NULL, y = NULL, tag = tag) + theme_fig10
}

fig10 <- wrap_plots(guild_panel("Herbivores", "a"), guild_panel("Invertivores", "b"),
                    guild_panel("Omnivores", "c"), guild_panel("Piscivores", "d"), ncol = 2)
fig10 <- wrap_elements(fig10) + labs(tag = "MaxN per deployment") +
  theme(plot.tag = element_text(size = 22, angle = 90), plot.tag.position = "left")
fig10
ggsave(file.path(fig_dir, "Fig10_maxn_by_guild.png"), fig10, width = 11, height = 10, dpi = 400)
```



# 4. Session info

```{r session}
sessionInfo()
```
